---
title: "Course Project 2: NOAA Severe Weather Events Analysis"
output: html_document
---
## Synopsis
### This project analyzes data provided by the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.  The data provides information about fatalities, injuries, and property and crop damage from various storm types.  
###In this project, I first identify the storm events that lead to the top 100 fatalities and injuries.  Next, I find the storm events that produce the top 100 property damage and crop damage costs, as well as the top 100 total damage costs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing
```{r cars, echo=TRUE}
loc <- "/Users/lilafakharzadeh/Desktop"
setwd(loc)
dat<-read.csv("repdata_data_StormData.csv",header=TRUE)
library(ggplot2)
library(dplyr)
library(plyr)
```


## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

## Top 100 Rankings by number of fatalities
```{r Fatalities, echo=TRUE}
countDatFat<-aggregate(dat$FATALITIES, by=list(dat$EVTYPE),FUN=sum)
orderCountDatFat<-head(countDatFat[order(-countDatFat$x),],100)
```

## Top 100 Rankings by number of Injuries
```{r Injuries, echo=TRUE}
countDatInj<-aggregate(dat$INJURIES, by=list(dat$EVTYPE),FUN=sum)
orderCountDatInj<-head(countDatInj[order(-countDatInj$x),],100)

```

## Merge fatality and injury rankings
```{r Merge, echo=TRUE}
orderCountDatFat$Type<-"FATALITIES"
orderCountDatInj$Type<-"INJURIES"
mergedFatInj<-rbind(orderCountDatFat,orderCountDatInj)
print(mergedFatInj)
```


## Across the United States, which types of events have the greatest economic consequences?

## Calculate property and crop damage by multiplying by respective factors
```{r Damage, echo=TRUE}
tPROPDMG<- mapvalues(dat$PROPDMGEXP, c("K","M","","B","m","+","0","5","6","?","4","2","3","h","7","H","-","1","8"), 
c(1e3,1e6,1,1e9,1e6,1,1,1e5,1e6,1,1e4,1e2,1e3,1,1e7,1e2,1,10,1e8))

tCROPDMG <- mapvalues(dat$CROPDMGEXP,
c("","M","K","m","B","?","0","k","2"),
c( 1,1e6,1e3,1e6,1e9,1,1,1e3,1e2))

dat$PROPDMGTOTAL<- as.numeric(tPROPDMG)*dat$PROPDMG
dat$CROPDMGTOTAL<- as.numeric(tCROPDMG)*dat$CROPDMG
```

## Calculate total damage
```{r TotalDamage, echo=TRUE}
dat$DMGTOTAL<-dat$PROPDMGTOTAL + dat$CROPDMGTOTAL
```

## Group damage by event type and calculate total
```{r Damage by Event, echo=TRUE}
propDat<-aggregate(dat$PROPDMGTOTAL, by= list(dat$EVTYPE), FUN=sum)
propDat<-propDat[order(-propDat$x),][1:100,]

cropDat<-aggregate(dat$CROPDMGTOTAL, by= list(dat$EVTYPE), FUN=sum)
cropDat<-cropDat[order(-cropDat$x),][1:100,]

propCropDat<-aggregate(dat$DMGTOTAL, by=list(dat$EVTYPE), FUN=sum)
propCropDat<- propCropDat[order(-propCropDat$x),][1:100,]
```

## Results
```{r Results Figures, echo=TRUE}
#Top 10 Storm Events that Produced The Most Fatalities
ggplot(orderCountDatFat[1:10,], aes(x=Group.1, y=x))+
geom_bar(stat="identity")+
theme(axis.text.x = element_text(angle = 90, hjust = 1))+
xlab("Event Type")+
ylab("Fatalities")+
ggtitle("Top 10 Fatalities by Event Type")

#Top 10 Storm Events that Produced The Most Injuries
ggplot(orderCountDatInj[1:10,], aes(x=Group.1, y=x))+
geom_bar(stat="identity")+
theme(axis.text.x = element_text(angle = 90, hjust = 1))+
xlab("Event Type")+
ylab("Injuries")+
ggtitle("Top 10 Injuries by Event Type")

#Top 10 Storm Events that Produced The Most Property/Crop Damage
ggplot(propCropDat[1:10,], aes(x=Group.1, y=x))+
geom_bar(stat="identity")+
theme(axis.text.x = element_text(angle = 90, hjust = 1))+
xlab("Event Type")+
ylab("Total Damage (Property + Crop)")+
ggtitle("Top Ten Event Types for Total Damage (Property + Crop)")
```




